Equivalence Testing and the Second Generation P-Value

### 2018-08-22 manuscript.Rmd
added abstract, expanded discussion

abstract: |
  We aim to compare TOST and SGPV.

keywords          : "equivalence testing, second generation p-values, p-values, hypothesis testing, statistics"
wordcount         : ""

@@

The TOST procedure also relies on the confidence interval around the effect. 
In the TOST procedure the data is tested against the lower equivalence bound in the first one-sided test, and against the upper equivalence bound in the second one-sided test. 
If both tests allow you to reject an effect as extreme or more extreme than the equivalence bound, you can reject the presence of an effect large enough to be meaningful, and conclude the observed effect is practically equivalent to zero. 
Because one-sided tests are performed, one can also conclude equivalence by checking whether the 1-2*alpha confidence interval (e.g., when the alpha level is 0.05, a 90% CI) falls completely within the equivalence bounds. 
Because both equivalence tests as the SGPV are based on whether and how much a confidence interval overlaps with equivalence bounds, it seems logical to compare the newly proposed SGPV to equivalence tests.

@@

To examine the relation between the TOST p-value and the SGPV we can calculate both statistics across a range of observed effect sizes. In \ref{fig:TOSTSGPV1} *p*-values are plotted for the TOST procedure and the SGPV. The statistics are calculated for hypothetical one-sample *t*-tests for *all* means that can be observed in studies ranging from 140 to 150 (on the x-axis). The equivalence range is set to 145 ± 2 (i.e., an equivalence range from 143 to 147), the observed standard deviation is assumed to be 2, and the sample size is 100. For example, for the left-most point in \ref{fig:TOSTSGPV1} the SGPV and the TOST *p*-value is calculated for a hypothetical study with a sample size of 100, an observed standard deviation of 2, and an observed mean of 140, where the *p*-value for the equivalence test is 1, and the SGPV is 0. Our conclusions about the relationship between TOST *p*-values and SGPV in this article are not dependent upon any specific example.

@@

The SGPV treats the equivalence range as the null-hypothesis, while the TOST procedure treats the values outside of the equivalence range as the null-hypothesis. For ease of comparison we can reverse the SGPV (by calculating 1-SGPV in Figure \\ref{fig:TOSTSGPV2}) to make the values more easily comparable. We see that the *p*-value from the TOST procedure and the SGPV follow each other closely.

@@

When the observed sample mean is 145 and we are testing against equivalence bounds of 143 and 147 using the TOST procedure for a one-sample *t*-test with a sample size of 30 and a standard deviation of 2, the equivalence test is significant, *t*(`r tost_res1$TOST_df`) = `r round(tost_res1$TOST_t1, digits = 2)`, *p* = `r round(tost_res1$TOST_p1, digits =  4)`. Because the 95% CI falls completely within the equivalence bounds, the SGPV is `r SGPV1` (see \ref{fig:TOSTSGPV1}).

On the other hand, if the observed mean is 140, the equivalence test is not significant (the observed mean is far outside the equivalence range of 143 to 147), *t*(`r tost_res2$TOST_df`) = `r round(tost_res2$TOST_t1, digits = 2)`, *p* = `r round(tost_res2$TOST_p1, digits =  4)` (or more accurately, *p* > .999 as *p*-values are bounded between 0 and 1). Because the 95% CI falls completely outside the equivalence bounds, the SGPV is `r SGPV2` (see \ref{fig:TOSTSGPV1}).

@@

It is clear the SGPV and the *p*-value from TOST are closely related. We can think of the SGPV as a straight line that will always overlap the *p*-value from an equivalence test in 3 points. When the TOST *p*-value is 0.5, the SGPV is also 0.5 (note that the reverse is not true). The SGPV is 50% when the observed mean falls exactly on the lower or upper equivalence bound. When the observed mean equals the equivalence bound, the difference between the mean in the data and the equivalence bound is 0, the *t*-value for the equivalence test is also 0, and thus the *p*-value is 0.5 (situation A, \ref{fig:TOSTSGPV3}). 
```{r, include=FALSE}
m <- 146.5
mu <- 144.5
sd <- 500
n <- 1000000
low_eqbound = -2 
high_eqbound = 2 
alpha = 0.025

@@

```
```{r, TOSTSGPV3, fig.cap="Means, normal distribution, and 95% CI for three example datasets that illustrate the relationship between *p*-values from TOST and SGPV."}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(-3, 5),
     yaxt = "n",
     ylab = "SGPV or TOST p-value",
     xlab = "Mean Difference")
axis(2, at = c(0.25,0.5,0.75), labels = c("C", "B", "A"), las = 1)
axis(1, at = c(-3,-2,-1,0,1,2,3,4,5), las = 1)
abline(v = tost_res1$high_eqbound, 
       lty = 2)
abline(v = tost_res1$low_eqbound, 
       lty = 2)
abline(v = 0, 
       lty = 2, 
       col = "grey")

@@

Two other points always have to overlap. When the 95% CI falls completely, but only just inside the equivalence region, the TOST (which relies on a one-sided test) should be significant at an alpha level of 0.025. When the SGPV changes from <1 to 1 the 95% CI is exactly equal to one of the equivalence bounds (see situation B in \ref{fig:TOSTSGPV3}, where the 95% CI falls completely inside the equivalence bounds) the TOST *p*-value is 0.025. The third point where the SGPV and the *p*-value from the TOST procedure should overlap is where the SGPV  changes from a positive value (i.e., 0.0001) to 0 (when the 95% CI completely falls outside of the equivalence bound, see situation C in \ref{fig:TOSTSGPV3}). When the 95% CI touches the outside of the equivalence bound and the TOST *p*-value will be 0.975.

@@

One weakness of the SGPV is its reliance on the 'small sample correction', where the SGPV is set to 0.5 whenever the ratio of the confidence interval width to the equivalence range exceeds 2:1 and the CI overlaps with the upper and lower bounds. This exception to the normal calculation of the SGPV is introduced to prevent misleading values. Without this correction it is possible that a confidence interval is extremely wide, and an equivalence range is extremely narrow, which without the correction would lead to a very low value for the SGPV, which as @blume_second-generation_2018 seem to suggest that 'the data favor alternative hypotheses', even when the data is just extremely inaccurate compared to the width of the equivalence range. Although setting the SGPV to 0.5 whenever the ratio of the confidence interval width to the equivalence range exceeds 2:1 is necessary to prevent misleading results, it leads to a range of situations where the SGPV is set to 0.5, while the *p*-value from the TOST procedure continues to differentiate (see \ref{fig:TOSTSGPV6}). An important benefit of equivalence tests is that is does not need such a correction to prevent misleading results.

@@

>>>>

abstract: |
  One way to move beyond the limitations of null-hypothesis tests statistical approaches have been developed where the observed data is compared against an range of values that are equivalent to the absence of a meaningful effect. We compare the behavior of the recently proposed second generation *p*-value [@blume_second-generation_2018] with equivalence testing using the Two-One-Sided Tests procedure [@schuirmann_comparison_1987]. We show the two approaches yield almost identical results under optimal circumstances, but under suboptimal circumstances (e.g., when the confidence interval is wider than the equivalence range, or with asymmetric confidence intervals) the second generation *p*-value becomes difficult to interpret as a descriptive statistic. The second generation is always interpretable in a dichotomous manner when it is 0 or 1, but this can easily be confirmed visually, and does not require calculation. Equivalence tests yield more consistent *p*-values and allow for a easy control of Type I and Type II error rates.

keywords          : "equivalence testing, second generation *p*-values, *p*-values, hypothesis testing, statistics"
wordcount         : ""

@@

The TOST procedure also relies on the confidence interval around the effect. 
In the TOST procedure the data is tested against the lower equivalence bound in the first one-sided test, and against the upper equivalence bound in the second one-sided test [@Lakens_equivalence_2018]. 
If both tests allow you to reject an effect as extreme or more extreme than the equivalence bound, you can reject the presence of an effect large enough to be meaningful, and conclude the observed effect is practically equivalent to zero. 
Because one-sided tests are performed, one can also conclude equivalence by checking whether the 1-2*alpha confidence interval (e.g., when the alpha level is 0.05, a 90% CI) falls completely within the equivalence bounds. 
Because both equivalence tests as the SGPV are based on whether and how much a confidence interval overlaps with equivalence bounds, it seems logical to compare the newly proposed SGPV to equivalence tests.

@@

The second generation *p*-value (SGPV) is calculated as:
$$
  p _ { \delta } = \frac { \left| I \cap H _ { 0 } \right| } { | I | } \times \max \left\{ \frac { | I | } { 2 \left| H _ { 0 } \right| } , 1 \right\}
$$
where I is the interval based on the data (e.g., a 95% confidence interval) and H~0~ is the equivalence range. The first term of this formula implies that the second generation *p*-value is the width of the confidence interval that overlaps with the equivalence range, divided by the total width of the confidence interval. The second term is a 'small sample correction' (see below) that comes into play whenever the confidence interval is more than twice as wide as the equivalence range.

To examine the relation between the TOST *p*-value and the SGPV we can calculate both statistics across a range of observed effect sizes. In \ref{fig:TOSTSGPV1} *p*-values are plotted for the TOST procedure and the SGPV. The statistics are calculated for hypothetical one-sample *t*-tests for *all* means that can be observed in studies ranging from 140 to 150 (on the x-axis). The equivalence range is set to 145 ± 2 (i.e., an equivalence range from 143 to 147), the observed standard deviation is assumed to be 2, and the sample size is 100. For example, for the left-most point in \ref{fig:TOSTSGPV1} the SGPV and the TOST *p*-value is calculated for a hypothetical study with a sample size of 100, an observed standard deviation of 2, and an observed mean of 140, where the *p*-value for the equivalence test is 1, and the SGPV is 0. Our conclusions about the relationship between TOST *p*-values and SGPV in this article are not dependent upon any specific example.

@@

The SGPV treats the equivalence range as the null-hypothesis, while the TOST procedure treats the values outside of the equivalence range as the null-hypothesis. For ease of comparison we can reverse the SGPV (by calculating 1-SGPV in Figure \ref{fig:TOSTSGPV2}) to make the values more easily comparable. We see that the *p*-value from the TOST procedure and the SGPV follow each other closely.

@@

When the observed sample mean is 145 and we are testing against equivalence bounds of 143 and 147 using the TOST procedure for a one-sample *t*-test with a sample size of 30 and a standard deviation of 2, the equivalence test is significant, *t*(`r tost_res1$TOST_df`) = `r round(tost_res1$TOST_t1, digits = 2)`, *p* = `r round(tost_res1$TOST_p1, digits =  5)`. Because the 95% CI falls completely within the equivalence bounds, the SGPV is `r SGPV1` (see Figure \ref{fig:TOSTSGPV1}).

On the other hand, if the observed mean is 140, the equivalence test is not significant (the observed mean is far outside the equivalence range of 143 to 147), *t*(`r tost_res2$TOST_df`) = `r round(tost_res2$TOST_t1, digits = 2)`, *p* = `r round(tost_res2$TOST_p1, digits =  4)` (or more accurately, *p* > .999 as *p*-values are bounded between 0 and 1). Because the 95% CI falls completely outside the equivalence bounds, the SGPV is `r SGPV2` (see Figure  \ref{fig:TOSTSGPV1}).

@@

It is clear the SGPV and the *p*-value from TOST are closely related. When confidence intervals are symmatric we can think of the SGPV as a straight line that is directly related to the *p*-value from an equivalence test for three values. When the TOST *p*-value is 0.5, the SGPV is also 0.5 (note that the reverse is not true). The SGPV is 50% when the observed mean falls exactly on the lower or upper equivalence bound. When the observed mean equals the equivalence bound, the difference between the mean in the data and the equivalence bound is 0, the *t*-value for the equivalence test is also 0, and thus the *p*-value is 0.5 (situation A, \ref{fig:TOSTSGPV3}).

```{r, include=FALSE}
m <- 146.5
mu <- 144.5
sd <- 500
n <- 1000000
low_eqbound = -2 
high_eqbound = 2 
alpha = 0.025

@@

```{r, TOSTSGPV3, fig.cap="Means, normal distribution, and 95% CI for three example datasets that illustrate the relationship between *p*-values from TOST and SGPV."}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(-3, 5),
     yaxt = "n",
     ylab = "SGPV or TOST p-value",
     xlab = "Mean Difference")
axis(2, at = c(0.25,0.5,0.75), labels = c("C", "B", "A"), las = 1)
axis(1, at = c(-3,-2,-1,0,1,2,3,4,5), las = 1)
abline(v = tost_res1$high_eqbound, 
       lty = 2)
abline(v = tost_res1$low_eqbound, 
       lty = 2)
abline(v = 0, 
       lty = 2, 
       col = "grey")

@@

Two other points always have to overlap. When the 95% CI falls completely, but only just inside the equivalence region, the TOST (which relies on a one-sided test) should be significant at an alpha level of 0.025. When the SGPV changes from <1 to 1 the 95% CI is exactly equal to one of the equivalence bounds (see situation B in \ref{fig:TOSTSGPV3}, where the 95% CI falls completely inside the equivalence bounds) the TOST *p*-value is 0.025. The third point where the SGPV and the *p*-value from the TOST procedure should overlap is where the SGPV  changes from a positive value to 0 (i.e., when the 95% CI falls completely outside of the equivalence bound, see situation C in \ref{fig:TOSTSGPV3}). When the 95% CI touches the outside of the equivalence bound and the TOST *p*-value will be 0.975.

@@

A final important weakness of the SGPV is its reliance on the 'small sample correction', where the SGPV is set to 0.5 whenever the ratio of the confidence interval width to the equivalence range exceeds 2:1 and the CI overlaps with the upper and lower bounds. This exception to the normal calculation of the SGPV is introduced to prevent misleading values. Without this correction it is possible that a confidence interval is extremely wide, and an equivalence range is extremely narrow, which without the correction would lead to a very low value for the SGPV, which as @blume_second-generation_2018 seem to suggest that 'the data favor alternative hypotheses', even when the data is just extremely inaccurate compared to the width of the equivalence range. Although setting the SGPV to 0.5 whenever the ratio of the confidence interval width to the equivalence range exceeds 2:1 is necessary to prevent misleading results, it leads to a range of situations where the SGPV is set to 0.5, while the *p*-value from the TOST procedure continues to differentiate (see \ref{fig:TOSTSGPV6}). An important benefit of equivalence tests is that is does not need such a correction to prevent misleading results.

@@

@blume_second-generation_2018 argue that the SGPV has improved eType I rror control. However, they mean that the SGPV has a *lower* error rate, not a *more accurate* error rate. In a Neyman-Pearson perspective (which forms the basis of equivalence tests) the goal is not to end up with an error rate that is as low as possible, but with a decision procedure that yields a desired error rate accurately. If a lower Type I error rate is desired, the alpha level can be lowered. Equivalence tests also make it straightforward to control the Type 1 error rate by performing an a-priori power analysis.
